From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-06 Pi+/paper/ufin.tex
working on ufin

The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are sets and
whose (iso)morphisms as bijections. The semantics interprets types as objects in $\BFin$, 1-combinators as isomorphisms
in $\BFin$, and for every pair of 1-combinators related by a 2-combinator, their interpretations are extensionally equal
in $\BFin$. This groupoid is strict: two equal permutations are extensionally identified, without an explicit witness
for the equality. In order to establish completeness, we want to quote back from the semantics to the
syntax. Specifically, given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted
1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$.

In a weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are
higher invertible cells for the left and right units of composition, and associativity. The higher cells come with their
own coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define
  2-category in appendix.} HoTT provides a rich internal language for working with weak higher groupoids, using the
``Types are weak $\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has
points for 0-cells, 1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible
1-cells.~\footnote{This is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the
  hom-categories are posets, that is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type
$\UFin$, which is a 1-groupoid. In this section, we develop the tools necessary to construct $\UFin$, starting with a
quick review of HoTT.

>>>>

The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
and whose (iso)morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
interpretations are extensionally equal in~$\BFin$. In other words, this groupoid is strict: two equal permutations are
extensionally identified, without an explicit witness for the equality. In order to establish completeness, we want to
quote back from the semantics to the syntax. Specifically, given a morphism in $\BFin$, we want to produce a
1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
witnessing the equality of the quoted 1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$
by replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}.

HoTT provides a rich internal language for working with weak higher groupoids, using the ``types are weak
$\infty$-groupoids'' correspondence. In this section, we show that the weakened version of the groupoid $\BFin$ can be
written in HoTT as a type $\UFin$, which is a 1-groupoid that has points for 0-cells, 1-paths for 1-cells, 2-paths for
2-cells with at most one 2-cell between compatible 1-cells.

%% In this section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.
%% Using the identity type for morphisms, a 1-groupoid in HoTT

IS THAT FOOTNOTE NECESSARY ??? We should avoid using footnotes as much as possible.\footnote{This is a locally-strict
  $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that is, truth-value
  enriched.}
